---
title: "Final Project: MLB Payroll vs Wins"
author: "Andrew Diveley"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
install.packages("googlesheets4")
library(googlesheets4)
```

## Overview

This project examines the relationship between team payroll and regular season wins in Major League Baseball from 2022 to 2024. Using data from Spotrac, I compiled regular season win totals and payroll figures for all 30 MLB teams across the three seasons.

The goal is to determine whether teams that spend more money tend to perform better during the regular season. I calculated descriptive statistics and ran linear regression models to explore the strength and consistency of the relationship between payroll and wins. I also included a visualization to highlight any patterns or outliers in the data.

The notebook ends with a story pitch that outlines the key findings, potential limitations, and a plan for how to report the results as a data journalism piece.

## Task 1: Data Exploration

My dataset includes regular season wins and team payrolls for all 30 MLB teams from 2022 to 2024. I got everything from Spotrac, which tracks how much each team spends on player salaries. Each team has win totals and payroll figures for each of the three seasons. The goal is to see if spending more money helps teams win more games.

The data only covers the regular season. It doesn’t include playoff performance or any player-level stats. That means I can’t look at how far teams go in the postseason or how individual contracts affect team results. Still, regular season wins give a clear picture of how teams perform over a long stretch.

The most important columns are the win totals and payroll amounts for each year. I’m looking to see if there’s a clear link between how much a team spends and how many games it wins. I’ll use those numbers to run regressions and make visualizations. Those will help me spot patterns or outliers.

One thing missing from the data is any context about why a team performed a certain way. I don’t have info on injuries, trades, or how well players played. I also don’t have advanced stats like run differential. So I can’t explain everything, but I can still look at big-picture trends.

I want to find out how strong the connection is between payroll and wins, and whether it changes from year to year. I also want to know if there are teams that keep doing better or worse than expected based on how much they spend. Finally, I want to see if there's a point where more spending stops helping as much.

```{r load-data}
mlb_data <- read_sheet("https://docs.google.com/spreadsheets/d/11bhOOC9QjHO-KDpsdtx7Veh8VaZYXTuSK1maateFKlI/edit?usp=sharing")

# Preview structure
View(mlb_data)
```
