# JOUR405: Statistics for Journalists
## Midterm Exam - Spring 2025

Name: Andrew Diveley

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences. Start by loading the tidyverse and any other libraries you think you might need.

```{r}
library(tidyverse)
```


## Part 1: Restaurant Health Inspections (15 points)

You want to understand how restaurants in Montgomery County are performing on health inspections. The first dataset contains restaurant health inspection scores for restaurants in Montgomery County. The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible and the compliance score and grade. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv` and complete these tasks:

### Tasks:
1. Calculate the mean and standard deviation of compliance scores (5 points)
2. Create a histogram of the compliance scores with a vertical line showing the mean (5 points)
3. Write 2-3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What would be newsworthy about this distribution? What's the story here? (5 points).

```{r}
library(tidyverse)
data <-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")

data_mean <- mean(data$compliance_score)
data_sd <- sd(data$compliance_score)

ggplot(data, aes(x = compliance_score)) +
  geom_histogram(binwidth = 0.5, fill = "red", color = "black") +
  geom_vline(xintercept = data_mean, color = "lightblue", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = data_mean + data_sd, color = "blue", linetype = "dotted", linewidth = 1) +
  geom_vline(xintercept = data_mean - data_sd, color = "blue", linetype = "dotted", linewidth = 1) +
  labs(title = "Compliance Scores with Mean and Standard Deviation",
       x = "Compliance Score",
       y = "Frequency")
```


The Distribution shows us that most people scored over 90%, with the standard deviation maxing out at 100% on the compliance score.  The data is very positively skewed.  If I were to write a story, I would single out the 4 stores that scored an F and possibly made comparisons to A-tier inspections.

## Part 2: High School Athletics (25 points)

You are reporting a story about high school sports participation in Maryland and want to see if there are differences between boys and girls. The second dataset shows participation numbers in high school sports across Maryland counties in 2024, broken down by sport and sex. Load the data from: `https://raw.githubusercontent.com/example/md_hs_sports_2024.csv` and complete these tasks:

### Tasks:
1. Calculate the correlation between boys' and girls' participation (5 points)
2. Add two columns called total and girls_pct using mutate(), with the total adding together boys and girls and girls_pct being the percentage of the total represented by girls participants. (5 points)
3. Create a scatterplot showing this relationship, adding a line of best fit (5 points)
4. In 2-3 sentences, explain what the correlation coefficient and scatterplot reveal about equity in Maryland high school sports participation. How do you interpret the school districts that are below the line vs those that are above? Which school districts are most worth examining further, and why? (10 points)

```{r}
library(tidyverse)
gender_data <-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_hs_participation.csv")

correlation <- cor(gender_data$boys, gender_data$girls)

gender_data <- gender_data %>%
  mutate(
    total = (boys + girls),
    girls_pct = (girls / total))

ggplot(gender_data, aes(x = total, y = girls_pct)) +
  geom_point() +
  labs(title = "girls vs. total",
       x = "total",
       y = "girls_pct")
```

The correlation coefficient of boys and girls if .982, meaning they have a strong positive relationsip, so there really was around the same amount of each gender in the study.  The scatterplot tells us that most schools hover around 40% to 60% of girls in sports (meaning there is high equity), but there are some notable dots.  The most obvious dots are Montgomery County (highest total but still has great gender diversity in sports), Baltimore County (middle total but has highest percentage of girls at 72%), and Somerset(low total and lowest girl percentage).

## Part 3: Public Transit Ridership (20 points)

You are investigating public transit ridership in the Washington, D.C. area and want to understand the patterns of daily bus and rail ridership. The third dataset contains daily bus and rail ridership totals from WMATA for the past year. Load the data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv and do the following:

### Tasks:
1. Calculate the average bus and rail ridership and standard deviation using summarize() (5 points)
2. Using the process you used in class, take a random sample daily ridership numbers and calculate the sample means and deviations for bus and rail. The number in the sample is up to you, but explain why you chose what you did. Compare this to the stats you generated in step 1. (5 points)
3. Using group_by() and summarize(), calculate the means for bus and rail ridership for each weekday. Describe the overall pattern of ridership for bus and rail - which days stand out and why? Are there differences between bus and rail in the standard deviation values? (10 points)


```{r}
library(tidyverse)
ridership_data <-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")

ridership_data %>%
  summarize(
    mean_total = mean(total),
    ridership_sd = sd(total))

set.seed(123)
sample10 <- sample_n(ridership_data, 10) ### There was a lot of data, so 5 felt like too little,

sample10_mean <- mean(sample10$total)
sample10_sd <- sd(sample10$total)

```

Some holidays have low ridership (Christmas and New Years), while some have high ridership (Halloween).  Another important thing to note is how Tues and Wed dominate high ridership days, while the weekend days tend to represent low ridership days.  This could most likely be because of work. On low ridership days, bus tends to be higher than rail, and on higher days, rail tends to be higher than bus.

## Part 4: Maryland Car Theft Rates (20 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to analyze the data to find out which counties have the highest rates. The fourth dataset contains car theft statistics for Maryland counties in 2023 and population. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv` and complete the following tasks:

### Tasks:
1. Using mutate, add a column that calculates the rate of car thefts for each county - you need to choose the per capita rate (5 points)
2. Calculate the median car theft rate and the total number of car thefts statewide. Which counties have rates above the median, and what percentage of all car thefts occur in those counties? (5 points)
3. Write 2-3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What's the lede of a story about your findings? (10 points)


```{r}
library(tidyverse)
thefts <-read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")

thefts <- thefts %>%
mutate(total_crime_rate = (X2023 / population) * 10000)

rate_median <- median(thefts$total_crime_rate)

```

There are 12 counties over the median, with Baltimore City easily leading the pack.  The distribution heavily lies on counties like Baltimore City and Prince George's County, because they are high population and high percentage.  The median is less than 1 person per 1000, and Baltimore City soars to 1 in 100, 10 times the median.  That would influence my lede:

"Baltimore City averages 10 times the median rate of car thefts" and then have a story about how they have changed over the years and what measures are being taken to prevent this.

## Part 5: Data Analysis Scenario (20 points)

You receive a tip that local emergency response times have gotten significantly worse over the past year. You obtain monthly data on response times for police, fire and ambulance calls.

Write 3-4 sentences (no code!) explaining:
1. What statistical measures would you calculate to verify this claim? (10 points)

I would most likely look at the mean response time and calculate the standard deviation of them all.

2. What visualizations would help readers understand the trends? (5 points)

I think people would like to look at a scatterplot of response time vs distance from police/fire department.  Or people could use a histogram of the average response time with the mean and standard deviation.

3. What additional context or data would you need to make this a complete story? (5 points)

There are so many factors, such as the emergency type, the distance from police/fire or whatever department, the amount of response needed (any ambulance needed, how many fire trucks and how many cop cars?).  There are so many variables.  I would like to use previous data as well, I can't just claim reponse times are going down.

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your analytical choices
- Before submitting, clear your environment and run the entire notebook

Good luck!
